Typing the Outputs

### 2015-08-21 sematics/boundary/boundary.tex
adding boundary type stuff

\newcommand {\ibksym} {\rotatebox[origin=c]{270}{$\therefore$}}
\newcommand {\btsym} {\rotatebox[origin=c]{270}{$\triangle$}}
\newcommand {\ktype} {\rm Type}
\newcommand {\ibk} [2] {{#1}~\ibksym~{#2}}
\newcommand {\btj} [4] {{#1}:{#2}~\myatsign~{#3}~\btsym~{#4}}
\newcommand {\btyo} [4] [\Gamma] {{#1}\vdash \btj{#2}{#3}\bbone{#4}}
\newcommand {\btyt} [4] [\Gamma] {{#1}\vdash \btj{#2}{#3}\bbtwo{#4}}
\newcommand {\btyw} [4] [\Gamma] {{#1}\vdash \btj{#2}{#3}w{#4}}
\newcommand {\btysub} [2] [] {\bty{e_{#1}:{#2}}{\tau_{#1}}{\sigma_{#1}}}

@@

For the most part, we'll do all of our work with a single new judgment \mbox{$\btj e A w \tau$}, 
which indicates that the expression $e$ (which has type $A$ at world $w$) has the boundary type $\tau$.
The goal is for the splitting rule to agree with the boundary type given to us by this judgment.
That is, if $\btj e A \bbtwo \tau$ and $\splittwo{e}{A}{p}{l}{r}$, 
then $p:\tau$ and $l:\tau \vdash r:A$ (and a similar relationship at stage \bbone, covered later).

\subsection{Straight-line Code}

Consider an input language which has no functions or divergence.
At either stage, the boundary type of any expression is 
just the product of the boundary types of its subexpressions
(unit in the nullary case).
The only exception to this pattern is the $\pause$ rule,
which also multiplies an integer into the boundary type.

\begin{mathpar}
\infer {\btyw i {\rm int}\rmunit} {\cdot} \and
\infer {\btyw b {\rm bool}\rmunit} {\cdot} \and
\infer {\btyw {\tup{e_1,e_2}} {A \times B} {\tau_1 \times \tau_2}} 
	{\btyw {e_1} A {\tau_1} 
	&\btyw {e_2} B {\tau_2}} \and
\infer {\btyw {\pio e} A \tau} 
	{\btyw {e} {A \times B} {\tau}} \and
\infer {\btyw {\pit e} B \tau} 
	{\btyw {e} {A \times B} {\tau}} \and
\infer {\btyo {\next e} {\fut A} \tau} 
	{\btyt {e} A {\tau}} \and
\infer {\btyt {\prev e} A \tau} 
	{\btyo {e} {\fut A} {\tau}} \and
\infer {\btyt {\pause e} \rmint {\tau \times \rmint}} 
	{\btyo e \rmint \tau}
\end{mathpar}

Finally, we look at the let rule, which introduces a variable into the context.\footnote{
Note the use of {\em hypothetical judgments}, 
wherein the judgements for a variable are given as assumptions at a declaration site rather than a rule for the reference site.
For now, the use of hypothetical judgments is merely a stylistic difference,
but it will become important when things get more complicated.}

\begin{mathpar}
\infer {\btyw {\letin x {e_1} {e_2}} {B} {\tau_1 \times \tau_2}} 
	{\btyw {e_1} A {\tau_1} 
	&\btyw [\Gamma,\btj x A w \rmunit] {e_2} B {\tau_2}}
\end{mathpar}

Notice how variable is introduced with a unit boundary value. 
This is because.

\section{First-Order Recursive Functions}

So to explore what this means, lets dial back the language a bit.  
In addition to straight-line from before, lets just add \texttt{if} expressions. 
It turns out this is easy to integrate into the existing framework:

\begin{mathpar}
\infer {\btyo {\ifthen {e_1}{e_2}{e_3}} {A} {\tau_1 \times (\tau_2 + \tau_3)}} 
	{\btyo {e_1} \rmbool {\tau_1} 
	&\btyo {e_2} A {\tau_2}
	&\btyo {e_3} A {\tau_3}} \and
\infer {\btyt {\ifthen {e_1}{e_2}{e_3}} {A} {\tau_1 \times \tau_2 \times \tau_3}} 
	{\btyt {e_1} \rmbool {\tau_1} 
	&\btyt {e_2} A {\tau_2}
	&\btyt {e_3} A {\tau_3}}
\end{mathpar}

The only particular care we took here was to observe speculative behavior in the stage \bbtwo\ case,
which induces a product (whereas stage \bbone\ had a sum).
We can rewrite the problematic example in our restricted language 
by distributing the application into the if and applying the lambda in both branches:
\[
\tallif {E}
	{\next{\pause{5}+1}}
	{\next{\pause{5}+\pause{5*5}}}
\]
We our new rules, we can say that the boundary type of this expression is
$\rmunit \times (\rmint + \rmint \times \rmint)$.

\section{Adding Functions and Fixed Points}

To get a taste of why the problem is hard,
we consider adding lots of features, namely \texttt{if}s, functions, and fixed points.
The astute reader will notice that these are enough to make the language Turing-complete.
Before getting into the rules, consider the following example, 
where $E$ is some purely stage \bbone\ expression:

\[
\left(
\tallif {E}
	{\lam x {\next{\pause{x}+1}}}
	{\lam x {\next{\pause{x}+\pause{x*x}}}}
\right ) 5
\]

What do we expect the eventual boundary value for this whole piece of code to be?
Well, it depends. 
If $E$ evaluates to true, then we'd expect the boundary value to be an integer (that is, have type $\rmint$),
but if $E$ evaluates to false, we'd expect the boundary value to be the tuple of two integers (that is, have type $\rmint \times \rmint$).
Ergo, the only way to have a boundary type for this whole piece of code is to either,
\begin{itemize}
\item figure out what $E$ evaluates, or
\item use a description for boundary types that mentions both $\rmint$ and $\rmint \times \rmint$, 
without committing to one.
\end{itemize}

Of course, since our language is Turing complete, figuring out what $E$ evaluates to is not decidable,
so we'll go with the second option.

\section{Higher-Order Functions}
\label{sec:higherorder}

Lets now take a different stab at this, by adding functions to our straight-line code,
but leaving \texttt{if}s out for now.
Again, lets consider an example:
\[
L1 := \lam x {\next{5}}
\]
The boundary type of $L1$ is obviously unit, right?
After all, lambdas are values, and values have unit boundary.
So then what about this example:
\[
L2 := \lam x {\next{\pause{x}+1}}
\]
I'd like to think it's the same story: unit boundary.
But what if we apply a value to these functions?
Consider
\[
\app {\lam x {\next{5}}} 3
\]
versus
\[
\app {\lam x {\next{\pause{x}+1}}} 3
\]
The boundary type of the first expression should still be unit 
(even after substitution it has no {\tt hold}s),
but the bound type of the second expression should definitely be int.

There's a problem here: $L1$ and $L2$ had the same boundary type before application, but different ones after.
This suggests that if we want the boundary type to be a compositional property,
then we'll need to keep track of more information than we do above.

\subsection{Adding Boundary Annotations}

Our first attempt at a solution will be to add more information to function types in our input language.
This new information simply tells us what the boundary type will be yielded when the function is applied.

[solution where we write the boundary type above the arrows]

\subsection{Separating Internal and External Boundaries}

There's something unsatisfying about this solution. 
The type system of the input language was perfectly functional before all this mess about boundary types.

[show how to separate the internal from external boundaries]

\subsection{Higher order funcitons}

[Take the system from the previous section and consider higher-order functions.]

% We do this by making a distinction between the {\em internal} boundary of an expression,
% and its {\em external} boundary.  
% $L1$ and $L2$ have the same external boundary, namely unit,
% but they have different internal boundaries.  
% In particular, the interal boundary of $L1$ should be int (or isomorphic to it), 
% whereas the internal boundary of $L2$ should be unit.

% Consider a language with products and functions.  The internal boundary kind is given by:

% \begin{mathpar}
% \infer {\ibk {\rm int}{\rm Unit}} {\cdot} \and 
% \infer {\ibk {\rm bool}{\rm Unit}} {\cdot} \and 
% \infer {\ibk {A \times B}{\kappa_A \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} \and 
% %\infer {\ibk {A \to B}{\kappa_A \to \ktype \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} 
% \end{mathpar}

>>>>

\newcommand {\ibksym} {\rotatebox[origin=c]{270}{$\therefore$}}
\newcommand {\btsym} {\rotatebox[origin=c]{270}{$\triangle$}}
\newcommand {\ibtsym} {\mathrel{\ooalign{$\btsym$\cr \hidewidth\hbox{$\bullet\mkern3mu$}\cr}}}
\newcommand {\ktype} {\rm Type}
\newcommand {\ibk} [2] {{#1}~\ibksym~{#2}}
\newcommand {\btj} [4] {{#1}:{#2}~\myatsign~{#3}~\btsym~{#4}}
\newcommand {\ibtj} [3] {{#1}\colfunsym{#2}~\myatsign~\bbone~\ibtsym~{#3}}
\newcommand {\btyo} [4] [\Gamma] {{#1}\vdash \btj{#2}{#3}\bbone{#4}}
\newcommand {\btyt} [4] [\Gamma] {{#1}\vdash \btj{#2}{#3}\bbtwo{#4}}
\newcommand {\btyw} [4] [\Gamma] {{#1}\vdash \btj{#2}{#3}w{#4}}
\newcommand {\btysub} [2] [] {\bty{e_{#1}:{#2}}{\tau_{#1}}{\sigma_{#1}}}
\newcommand {\ibtyo} [4] [\Gamma] {{#1}\vdash \ibtj{#2}{#3}{#4}}
\newcommand {\ibtyt} [4] [\Gamma] {{#1}\vdash \ibtj{#2}{#3}\bbtwo{#4}}

@@

In this section, we develop a boundary-type system for first-order recursive functions
with {\tt if}s, but no sum types.

\subsection {Core Features and {\tt if}s}

For the most part, we'll do all of our work with a single new judgment \mbox{$\btj e A w \tau$}, 
which indicates that the term $e$ (where $\colwor e A$ and $e \comp$) has the boundary type $\tau$.
The goal is for the splitting rule to agree with the boundary type given to us by this judgment.
That is, if $\btj e A \bbtwo \tau$ and $\splittwo{e}{A}{p}{l}{r}$, 
then $p:\tau$ and $l:\tau \vdash r:A$ (and a similar relationship at stage \bbone, covered later).

For core features at either stage, the boundary type of any term is 
just the product of the boundary types of its computation subterms.
The only exception to this pattern is the $\pause$ rule,
which also multiplies an integer into the boundary type.

\begin{framed}
\noindent\textbf{Core Feature Boundary Types:}
\begin{mathpar}
\infer {\btyw {\exv v} A\rmunit} {\cdot} \and
\infer {\btyw {\tup{e_1,e_2}} {A \times B} {\tau_1 \times \tau_2}} 
	{\btyw {e_1} A {\tau_1} 
	&\btyw {e_2} B {\tau_2}} \and
\infer {\btyw {\pio e} A \tau} 
	{\btyw {e} {A \times B} {\tau}} \and
\infer {\btyw {\pit e} B \tau} 
	{\btyw {e} {A \times B} {\tau}} \and
\infer {\btyo {\next e} {\fut A} \tau} 
	{\btyt {e} A {\tau}} \and
\infer {\btyt {\prev e} A \tau} 
	{\btyo {e} {\fut A} {\tau}} \and
\infer {\btyt {\pause e} \rmint {\tau \times \rmint}} 
	{\btyo e \rmint \tau} \and
\infer {\btyw {\letin x {e_1} {e_2}} {B} {\tau_1 \times \tau_2}} 
	{\btyw {e_1} A {\tau_1} 
	&\btyw {e_2} B {\tau_2}}
\end{mathpar}
\end{framed}

Note that in the {\tt let} rule, we needn't declare a boundary type for the $x$ variable,
since variables are values, and only computations have a boundary type.
Rules for \texttt{if} terms are similarly easy:

\begin{framed}
\noindent\textbf{If Expression Boundary Types:}
\begin{mathpar}
\infer {\btyo {\ifthen {e_1}{e_2}{e_3}} {A} {\tau_1 \times (\tau_2 + \tau_3)}} 
	{\btyo {e_1} \rmbool {\tau_1} 
	&\btyo {e_2} A {\tau_2}
	&\btyo {e_3} A {\tau_3}} \and
\infer {\btyt {\ifthen {e_1}{e_2}{e_3}} {A} {\tau_1 \times \tau_2 \times \tau_3}} 
	{\btyt {e_1} \rmbool {\tau_1} 
	&\btyt {e_2} A {\tau_2}
	&\btyt {e_3} A {\tau_3}}
\end{mathpar}
\end{framed}

The only particular care we took here was to observe speculative behavior in the stage \bbtwo\ case,
which induces a product (whereas stage \bbone\ had a sum).

\subsection {Functions}

Consider the following example, which is open on the variable \mbox{$\colfun g {\rmint \to \rmint} \bbone$}.
\[
\app g {\app g 5}
\]
Clearly, the boundary type of this computation depends on the particular function $g$ is bound to.
Thus, in order for a boundary type system to be complete,
the variable $g$ must be associated with information about the boundary type that is produced where $g$ is applied.
This is called the \emph{internal boundary type} of $g$.

Our system works by associating every function derivation \mbox{$\colfun g {A \to B} \bbone$} with an internal boundary type $\sigma$
via the judgement $\ibtj g {A \to B} \sigma$.  
Functions delcared at world $\bbtwo$ do not need internal boundar types.

\begin{framed}
\noindent\textbf{First-Order Recursive Function Boundary Types:}
\begin{mathpar}
\infer {\btyo {\app {f}{e}} {A} {\tau \times \sigma}} 
	{\ibtyo {f} {A \to B} {\sigma} 
	&\btyo {e} A {\tau}} \and
% \infer {\btyt {\fix fxe} {A} {\tau}} 
% 	{\btyt {e} A {\tau}} \and
% \infer {\btyt {\app {f}{e}} {A} {\tau}} 
% 	{\btyt {e} A {\tau}} \and
% \infer {\btyw {\letin x {e_1} {e_2}} {B} {\tau_1 \times \tau_2}} 
% 	{\btyw {e_1} A {\tau_1} 
% 	&\btyw {e_2} B {\tau_2}}
\end{mathpar}
\end{framed}

% \section{Adding Functions and Fixed Points}

% To get a taste of why the problem is hard,
% we consider adding lots of features, namely \texttt{if}s, functions, and fixed points.
% The astute reader will notice that these are enough to make the language Turing-complete.
% Before getting into the rules, consider the following example, 
% where $E$ is some purely stage \bbone\ expression:

% \[
% \left(
% \tallif {E}
% 	{\lam x {\next{\pause{x}+1}}}
% 	{\lam x {\next{\pause{x}+\pause{x*x}}}}
% \right ) 5
% \]

% What do we expect the eventual boundary value for this whole piece of code to be?
% Well, it depends. 
% If $E$ evaluates to true, then we'd expect the boundary value to be an integer (that is, have type $\rmint$),
% but if $E$ evaluates to false, we'd expect the boundary value to be the tuple of two integers (that is, have type $\rmint \times \rmint$).
% Ergo, the only way to have a boundary type for this whole piece of code is to either,
% \begin{itemize}
% \item figure out what $E$ evaluates, or
% \item use a description for boundary types that mentions both $\rmint$ and $\rmint \times \rmint$, 
% without committing to one.
% \end{itemize}

% Of course, since our language is Turing complete, figuring out what $E$ evaluates to is not decidable,
% so we'll go with the second option.

% \section{Higher-Order Functions}
% \label{sec:higherorder}

% Lets now take a different stab at this, by adding functions to our straight-line code,
% but leaving \texttt{if}s out for now.
% Again, lets consider an example:
% \[
% L1 := \lam x {\next{5}}
% \]
% The boundary type of $L1$ is obviously unit, right?
% After all, lambdas are values, and values have unit boundary.
% So then what about this example:
% \[
% L2 := \lam x {\next{\pause{x}+1}}
% \]
% I'd like to think it's the same story: unit boundary.
% But what if we apply a value to these functions?
% Consider
% \[
% \app {\lam x {\next{5}}} 3
% \]
% versus
% \[
% \app {\lam x {\next{\pause{x}+1}}} 3
% \]
% The boundary type of the first expression should still be unit 
% (even after substitution it has no {\tt hold}s),
% but the bound type of the second expression should definitely be int.

% There's a problem here: $L1$ and $L2$ had the same boundary type before application, but different ones after.
% This suggests that if we want the boundary type to be a compositional property,
% then we'll need to keep track of more information than we do above.

% \subsection{Adding Boundary Annotations}

% Our first attempt at a solution will be to add more information to function types in our input language.
% This new information simply tells us what the boundary type will be yielded when the function is applied.

% [solution where we write the boundary type above the arrows]

% \subsection{Separating Internal and External Boundaries}

% There's something unsatisfying about this solution. 
% The type system of the input language was perfectly functional before all this mess about boundary types.

% [show how to separate the internal from external boundaries]

% \subsection{Higher order funcitons}

% [Take the system from the previous section and consider higher-order functions.]

% % We do this by making a distinction between the {\em internal} boundary of an expression,
% % and its {\em external} boundary.  
% % $L1$ and $L2$ have the same external boundary, namely unit,
% % but they have different internal boundaries.  
% % In particular, the interal boundary of $L1$ should be int (or isomorphic to it), 
% % whereas the internal boundary of $L2$ should be unit.

% % Consider a language with products and functions.  The internal boundary kind is given by:

% % \begin{mathpar}
% % \infer {\ibk {\rm int}{\rm Unit}} {\cdot} \and 
% % \infer {\ibk {\rm bool}{\rm Unit}} {\cdot} \and 
% % \infer {\ibk {A \times B}{\kappa_A \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} \and 
% % %\infer {\ibk {A \to B}{\kappa_A \to \ktype \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} 
% % \end{mathpar}
